---
title: "Maven NYC Taxi Challenge"
author: "Abe"
date: "10/17/2021"
output: pdf_document
---
Calling packages
Using the tidyverse because it has the libraries I need to clean, mutate and visualize data.
library(data.table) - helps load data faster then read csv
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(dplyr)
library(magrittr)
```

Loading all of the excel files into Rstudio
```{r}
taxi_data_2017 <- fread("C:/Users/abedi/OneDrive/Desktop/NYC_Taxi_Trip R_project/taxi_trips/2017_taxi_trips.csv")
taxi_data_2018 <- fread("C:/Users/abedi/OneDrive/Desktop/NYC_Taxi_Trip R_project/taxi_trips/2018_taxi_trips.csv")
```


```{r}
taxi_data_2019 <- fread("C:/Users/abedi/OneDrive/Desktop/NYC_Taxi_Trip R_project/taxi_trips/2019_taxi_trips.csv")
taxi_data_2020 <- fread("C:/Users/abedi/OneDrive/Desktop/NYC_Taxi_Trip R_project/taxi_trips/2020_taxi_trips.csv")
```

Noticed taxi_data_2019 & 2020 have 19 columns checked the first 6 rows to see what it is. 
```{r}
head(taxi_data_2019)
```
```{r}
head(taxi_data_2020)
```
Congestion_surcharge is the extra column which will prevent the union_all from working because all of the columns need to be the same. Going to remove the extra columns for both 2019 & 2020.
```{r}
taxi_data_2019$congestion_surcharge <- NULL
```

```{r}
taxi_data_2020$congestion_surcharge <- NULL
```

**Did not remove any columns as previously stated because removing them caused an issued with the union_all.**

At first I tried to union_all the 4 dataframes which took too long which resulted in the total rows being 8M less. I decided to conduct two separate union_all statements to fix the issue. 
```{r}
taxi_data_all <- union_all(taxi_data_2017, taxi_data_2018)
```

```{r}
taxi_data_all_2 <- union_all(taxi_data_2019, taxi_data_2020)
```

```{r}
taxi_data_all_final <- union_all(taxi_data_all, taxi_data_all_2)
```


Garbage collection function to print memory usage statistics as previously I was receiving this error 
**Error: cannot allocate vector of size 216.1 Mb**
```{r}
gc()
```

Checked memory limit and noticed it was 12049
```{r}
memory.limit()
```

Increased limit as noted below and ran the final union_all again and summed up the dataframes to compare the final union_all to confirm 28M rows were unioned successfully.
```{r}
memory.limit(size = 50000)
```

Removed previous dataframes to save GB in Global Environment
```{r}
remove(taxi_data_2017, taxi_data_2018, taxi_data_2019, taxi_data_2020, taxi_data_all, taxi_data_all_2)
```


View data types
```{r}
glimpse(taxi_data_all_final)
```

Cleaning requirements:

1. Remove trips that were NOT sent via "store and forward".
```{r}
taxi_data_clean <- taxi_data_all_final %>% 
  filter(store_and_fwd_flag == "N")
```

```{r}
remove(taxi_data_all_final)
```


2. Only keep street-hailed trips paid by card or cash with a standard rate.
```{r}
taxi_data_clean <- taxi_data_clean  %>% 
  filter(RatecodeID == 1 & (payment_type < 3))
#RateCodeID & payment_type are both int
```

```{r}
head(taxi_data_clean)
```

3. Remove any trips before 2017 or after 2020, along with any trips or pickups or drop-offs into unknown zones.
```{r}
taxi_data_clean <- taxi_data_clean  %>% 
  filter(between (lpep_pickup_datetime, as.Date("2017-01-01"), as.Date ("2021-01-01")))
```

4. Any trips with no recorded passengers had 1 passenger.
```{r}
taxi_data_clean <- taxi_data_clean  %>% 
mutate(passenger_count = if_else(passenger_count == 0, 1, 1)) 
```

Verify it worked ^
```{r}
min(taxi_data_clean$passenger_count)
```
5. If pickup date/time is AFTER drop-off date/time, swap them.
```{r}
taxi_data_clean <- taxi_data_clean  %>% 
mutate(lpep_pickup_datetime = if_else(lpep_dropoff_datetime < lpep_pickup_datetime, lpep_dropoff_datetime, lpep_pickup_datetime))
```

Cleaning step worked ^
```{r}
taxi_data_clean %>% 
filter(lpep_dropoff_datetime < lpep_pickup_datetime)
```


6. Remove any trips lasting longer than a day, and any trips which show a distance and fare amount of zero.
```{r}
taxi_data_clean <- taxi_data_clean %>% 
  mutate(trip_hours = difftime(taxi_data_clean$lpep_dropoff_datetime, taxi_data_clean$lpep_pickup_datetime, units = "hours"))
```


Verifying the time difference
```{r}
max(taxi_data_clean$trip_hours)
```

```{r}
taxi_data_clean <- taxi_data_clean  %>% 
  filter(trip_hours <= 24) %>% 
  filter(trip_distance > 0 & (fare_amount > 0))
```

Verifying that anything greater then a day and removed anything that was 0. 
```{r}
max(taxi_data_clean$trip_hours)
min(taxi_data_clean$trip_distance)
min(taxi_data_clean$fare_amount)
```


7. If any fare, taxed and surcharge are negative make positive.
```{r}
min(taxi_data_clean$fare_amount)
min(taxi_data_clean$mta_tax)
min(taxi_data_clean$congestion_surcharge)
summary(taxi_data_clean$congestion_surcharge)
```
Removing columns that are not required for analysis
```{r}
taxi_data_clean$VendorID <- NULL
taxi_data_clean$RatecodeID <- NULL
```
Another way to remove a column by name
```{r}
select(taxi_data_clean, -extra, -mta_tax, -tolls_amount)
```
Importing the next dataframe
```{r}
taxi_zones <- fread("C:/Users/abedi/OneDrive/Desktop/NYC_Taxi_Trip R_project/taxi_zones.csv")
```

Viewing the new dataframe
```{r}
head(taxi_zones)
```

Memory limit was too small had to increase again
```{r}
memory.limit(size = 100000)
```


Joining the new dataframe by pickup location ID
```{r}
taxi_data_join <- taxi_data_clean  %>% 
  inner_join(taxi_zones, by = c("PULocationID" = "LocationID"))
#View(taxi_data_join)#
```

Joining the new dataframe by dropoff location ID
```{r}
taxi_data_join <- taxi_data_join  %>% 
  inner_join(taxi_zones, by = c("DOLocationID" = "LocationID"))
```


Renaming columns
```{r}
taxi_data_join <- taxi_data_join %>% 
  #select(Borough.x, Zone.x, service_zone.x, Borough.y, Zone.y, service_zone.y) %>% 
  rename(PU_borough = Borough.x , PU_zone = Zone.x, PU_service_zone = service_zone.x, DO_borough = Borough.y , DO_zone = Zone.y, DO_service_zone = service_zone.y)
head(taxi_data_join)
```

Returning Weekday
```{r}
taxi_data_join$weekday = wday(taxi_data_join$lpep_pickup_datetime, label = TRUE, abbr = FALSE)
```

Returning Day of Week
```{r}
taxi_data_join$dow = wday(taxi_data_join$lpep_pickup_datetime, label = FALSE, week_start = getOption("lubridate.week.start", 7))
```

Returning Day of month
```{r}
taxi_data_join$month_day = mday(taxi_data_join$lpep_pickup_datetime)
```

Extracting year from date/time column
```{r}
taxi_data_join$year <- year(taxi_data_join$lpep_pickup_datetime)
```

Extracting month name
```{r}
taxi_data_join$month_name = month(taxi_data_join$lpep_pick, label = TRUE, abbr = FALSE)
```

Extracting date/time
```{r}
taxi_data_join_2 <-taxi_data_join %>% 
  mutate(
    date = as.Date(taxi_data_join$lpep_pickup_datetime),
    hour = hour(taxi_data_join$lpep_pickup_datetime),
    minute = minute(taxi_data_join$lpep_pickup_datetime),
    second = second(taxi_data_join$lpep_pickup_datetime)
    #)  %>% 
 # mutate(
    #service_date = format(date,"%m/%d/%Y"),
    #PU_time = paste(hour, minute, second, sep = ":")
  #) %>% 
 # mutate( 
      #hour = hour(taxi_data_join$lpep_dropoff_datetime),
      #minute = minute(taxi_data_join$lpep_dropoff_datetime),
      #second = second(taxi_data_join$lpep_dropoff_datetime)
  #) %>% 
  #mutate(
    #DO_time = paste(hour, minute, second, sep = ":")
  )
# During the visualization part of this project I noticed the date and times previously extracted transformed into characters which cannot be visualized.
```

Removing more columns 
```{r}
select(taxi_data_join_2, -minute, -second)
```

Relocating the join result
```{r}
taxi_data_join_3 <- taxi_data_join_2 %>% 
  relocate(PU_borough, PU_zone, PU_service_zone, .before = DOLocationID) %>% 
  relocate(DO_borough, DO_zone, DO_service_zone, .before = passenger_count) %>% 
  relocate(store_and_fwd_flag, .before = passenger_count) %>% 
  relocate(date, .before = PULocationID)
```

This didn't work for some reason to remove columns
```{r}
taxi_data_join_4 <- taxi_data_join_3  
select(taxi_data_join_4, -minute, -second)
```

Removed columns using the below method again.
```{r}
taxi_data_join_4$minute <- NULL
taxi_data_join_4$second <- NULL
```

```{r}
head(taxi_data_join_4)
```

```{r}
remove(taxi_data_clean, taxi_data_join, taxi_data_join_2, taxi_data_join_3)
```


Saving the final clean copy of data to use in separate r markdown. Pedro recommends always have 2 markdowns one for cleaning and one for visualizing. 
```{r}
fwrite(taxi_data_join_4, "C:/Users/abedi/OneDrive/Documents/R Projects/Maven Analytics NYC Taxi Challenge/clean_data_3.csv")
```

